Evaluating infectious disease forecasts with allocation scoring rules

### 2023-08-31 alloscore_manuscript/allocation.tex

In the decision making setting that we consider, an action $\mathbf{x} = (x_1, \ldots, x_N)$ is a vector specifying the amount that is allocated to each of $N$ locations. We require that each $x_i$ is non-negative and that the total allocation across all locations equals the amount of available resources, $K$: $\sum_{i=1}^N x_i = K$. The set $\mathcal{X}$ consists of all possible allocations that satisfy these constraints. The eventually realized resource need in each location is denoted by $\mathbf{y} = (y_1, \ldots, y_N)$. These levels of need are not known at the time of decision making, so we define the random vector $Y = (Y_1, \ldots, Y_N)$ where $Y_i$ represents the as-yet-unknown level of resource need in location $i$. Forecasts of need in each location are collected in $F = (F_1, \ldots, F_N)$. We assume that resource need is non-negative and the forecasts reflect that, i.e. the support of each $F_i$ is a subset of $\mathbb{R}^+$. Finally, we assume that each unit of unmet need incurs a loss denoted by $L$, so that if the selected resource level $x_i$ in location $i$ is less than the realized need $y_i$, a loss of $L(y_i - x_i)$ results.

@@

\paragraph{Step 1: specify a loss function.} The loss associated with a particular allocation is calculated by summing contributions from unmet need in each location:
\begin{equation}
s_A(\mathbf{x}, \mathbf{y}) = \sum_{i=1}^N L (y_i - x_i)_+. \label{eqn:loss_fn}
\end{equation}
Here, $(y_i - x_i)_+$ is the unmet need in location $i$, which is given by $y_i - x_i$ if the realized need $y_i$ in location $i$ is greater than the amount $x_i$ allocated to that location, or $0$ if the amount $x_i$ allocated to unit $i$ is greater than or equal to the realized need.

\paragraph{Step 2: Given a probabilistic forecast $F$, identify the Bayes act.} The Bayes act associated with the forecast, $\mathbf{x}^F$, is the allocation that minimizes the expected loss:
\begin{align}
    \mathbf{x}^F &= \mathrm{arg min}_{\mathbf{x}}\,\, \mathbb{E}_{F} [s_A(\mathbf{x}, \mathbf{Y})] \text{ subject to }
    0 \leq x_i \text{ for all $i$ and} \, \sum_{i=1}^N x_i = K, \label{AP}
\end{align}
where $\mathbb{E}_{F} [s_A(\mathbf{x}, \mathbf{Y})] = \sum_{i=1}^{N} L \mathbb{E}_{F_i}[(Y_i - x_i)_+]$ sums the expected loss due to unmet need across all locations. It can be shown that the Bayes act has $x^F_i = F_i^{-1}(1 - \lambda^{\star}/L)$, where $\lambda^{\star}$ depends on the problem parameters $K$ and $L$ as well as the forecast distributions and is chosen so as to satisfy the equation
\begin{align}
\sum_{i=1}^{N}F_i^{-1}(1 - \lambda^{\star}/L) = K.
\end{align}
This partial solution to the allocation problem seems to have first appeared in \cite{hadleywhitin1963}; see the supplemental materials for a derivation.

One interpretation of this result is that the Bayes act sets the allocation in each location $i$ to a quantile of the forecast distribution $F_i$ for that location. The quantile is at a probability level $\alpha = 1 - \lambda^{\star}/L$ that is the same for all locations, and is chosen such that the constraint is satisfied. According to this interpretation, we divide the available resources among the locations in such a way that the forecasted probability that need will exceed the available resources is the same in all locations.
An alternative interpretation comes from noting that for each location $i$, $\frac{\partial}{\partial x_i} \mathbb{E}_{F} [s_A(\mathbf{x}, \mathbf{Y})] \vert_{\mathbf{x}=\mathbf{x}^F} = - \lambda^{\star}$ (see the supplement for a proof).
In words, at the allocation given by the Bayes act, the rate of change of the expected score as a function of the amount allocated to location $i$ is given by $-\lambda^{\star}$.
This derivative is the same for all locations, so the optimal allocation divides the available resources across all locations in such a way that according to $F$, the expected benefit of 1 additional unit of resources is the same in all locations.

\paragraph{Step 3: Define the scoring rule.} We can now define a proper scoring rule for the probabilistic forecast $F$ as
\begin{equation}
S_A(F, y; K) = s_A(\mathbf{x}^F, y) = \sum_{i=1}^n L (y_i - F_i^{-1}(1 - \lambda^{\star}/L))_+
\end{equation}
This score measures the total unmet need across all locations that results from using the Bayes allocation associated with the forecast $F$ when the actual level of need in each location is observed to be $y_i$.

>>>>

In the decision making setting that we consider, an action $x = (x_1, \ldots, x_N)$ is a vector specifying the amount that is allocated to each of $N$ locations. We require that $0 \leq x$, i.e., that each $x_i$ is non-negative, and that the total allocation across all locations equals the amount of available resources, $K$: $\sum_{i=1}^N x_i = K$. The set $\mathcal{X}$ consists of all possible allocations that satisfy these constraints. The eventually realized resource need in each location is denoted by $y = (y_1, \ldots, y_N)$. These levels of need are not known at the time of decision making, so we define the random vector $Y = (Y_1, \ldots, Y_N)$ where $Y_i$ represents the as-yet-unknown level of resource need in location $i$. Forecasts of need in each location are collected in $F = (F_1, \ldots, F_N)$. We assume that resource need is non-negative and the forecasts reflect that, i.e. the support of each $F_i$ is a subset of $\mathbb{R}^+$. Finally, we assume that each unit of unmet need incurs a loss denoted by $L$, so that if the selected resource level $x_i$ in location $i$ is less than the realized need $y_i$, a loss of $L(y_i - x_i)$ results.

@@

\paragraph{Step 1: specify a loss function.} The loss associated with a particular allocation is calculated by summing contributions from unmet need in each location:
\begin{equation}
s_A(x, y) = \sum_{i=1}^N L (y_i - x_i)_+. \label{eqn:loss_fn}
\end{equation}
Here, $(y_i - x_i)_+$ is the unmet need in location $i$, which is given by $y_i - x_i$ if the realized need $y_i$ in location $i$ is greater than the amount $x_i$ allocated to that location, or $0$ if the amount $x_i$ allocated to unit $i$ is greater than or equal to the realized need.

\paragraph{Step 2: Given a probabilistic forecast $F$, identify the Bayes act.} The Bayes act associated with the forecast, $x^F$, is the allocation that minimizes the expected loss:
\begin{align}
    x^F &= \underset{0 \leq x}{\mathrm{arg min}}\,\, \mathbb{E}_{F} [s_A(x, Y)] \text{ subject to }
     \, \sum_{i=1}^N x_i = K, \label{AP}
\end{align}
where $\mathbb{E}_{F} [s_A(x, Y)] = \sum_{i=1}^{N} L \mathbb{E}_{F_i}[(Y_i - x_i)_+]$ sums the expected loss due to unmet need across all locations. It can be shown that the Bayes act has $x^F_i = F_i^{-1}(1 - \lambda^{\star}/L)$, where $\lambda^{\star}$ depends on the problem parameters $K$ and $L$ as well as the forecast distributions and is chosen so as to satisfy the equation
\begin{align}
\sum_{i=1}^{N}F_i^{-1}(1 - \lambda^{\star}/L) = K.
\end{align}
This partial solution to the allocation problem seems to have first appeared in \cite{hadleywhitin1963}; see the supplemental materials for a derivation.

One interpretation of this result is that the Bayes act sets the allocation in each location $i$ to a quantile of the forecast distribution $F_i$ for that location. The quantile is at a probability level $\alpha = 1 - \lambda^{\star}/L$ that is the same for all locations, and is chosen such that the constraint is satisfied. According to this interpretation, we divide the available resources among the locations in such a way that the forecasted probability that need will exceed the available resources is the same in all locations.
An alternative interpretation comes from noting that for each location $i$, $\frac{\partial}{\partial x_i} \mathbb{E}_{F} [s_A(x, Y)] \vert_{x=x^F} = - \lambda^{\star}$ (see the supplement for a proof).
In words, at the allocation given by the Bayes act, the rate of change of the expected score as a function of the amount allocated to location $i$ is given by $-\lambda^{\star}$.
This derivative is the same for all locations, so the optimal allocation divides the available resources across all locations in such a way that according to $F$, the expected benefit of 1 additional unit of resources is the same in all locations.

\paragraph{Step 3: Define the scoring rule.} We can now define a proper scoring rule for the probabilistic forecast $F$ as
\begin{equation}
S_A(F, y; K) = s_A(x^F, y) = \sum_{i=1}^n L (y_i - F_i^{-1}(1 - \lambda^{\star}/L))_+
\end{equation}
This score measures the total unmet need across all locations that results from using the Bayes allocation associated with the forecast $F$ when the actual level of need in each location is observed to be $y_i$.
